\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed

### 2014-10-06 Paper/grass.tex
Almost finished language edit

\documentclass[aps,prl,twocolumn,superscriptaddress,10pt]{revtex4-1}  % for review and submission
%\documentclass[aps,twocolumn,floatfix,prl,10pt]{revtex4-1}
%\documentclass[letterpaper,10pt,prl,twocolumn,aps]{revtex4-1}
%\documentclass[aps,preprint,floatfix,prl]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color}
\usepackage{longtable}
\usepackage{array}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\graphicspath{ {images/}{images/Growth_Rate_Cont_Plots/} }
\newcommand{\bn}{{\boldsymbol{\hat{n}}}}
\newcommand{\bt}{{\boldsymbol{\hat{t}}}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand{\del}{\partial}
\newcommand{\hg}{h_g}
\newcommand{\Rey}{\text{R}}
\newcommand{\Ndg}{\tilde{N}_g}
\newcommand{\monami}{\textit{monami}}
\newcommand{\ubl}{u_\text{bl}}
\definecolor{tableShade}{gray}{0.8}

\begin{document}
\title{\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
\author{L. Mahadevan}
\affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi}
\affiliation{OIST Graduate University, Okinawa 904-0495, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hole Oceanographic Institution, Woods Hole MA 02543 USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

Therefore, to develop the essential mathematical model, we assume the vegetation to be rigid, and verify it \textit{a posteriori}.

@@

The former arises because, in the case of dense vegetation, the shear stress exerted by the bottom surface is expected to be negligible compared to the vegetation drag. 

@@

Because this boundary layer develops from purely local dynamics, independent of the influence of the boundaries, and we identify to be analogous to the free shear layer \cite{Ghisal02} in the previous explanation of \monami.

@@

Solving for $\delta$ yields $\delta=(\Rey\tilde{N_g})^{-1/3}H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, and $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow. 

@@

The asymptotic regime of a thin boundary layer is expected to hold for $\Rey \Ndg \gtrsim 100$.

@@

To understand the evolution of small perturbations to the steady flow profile, we substitute $\bu = (U+\tilde{u}, \tilde{v})$, $p=P+\tilde{p}$ in eq.~\eqref{navier-stokes} and expand to linear order to obtain (after dropping the tilde decoration)

@@

\begin{table*}
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.4}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{Equation \eqref{base_equ}} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& $\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Dense grass limit &  & Equation \eqref{eqn:mode1asymp} & Equation \eqref{eqn:mode2asymp}  \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{1/2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto H/\delta$		& $\propto H/\delta$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& ~~yes, near $y=\hg$~~			& ~~no, spans water column~~
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the mechanism of waving and the origin of the threshold, we consider the behavior of the instability as a function of the vegetation drag characterized by $\Ndg$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical vegetation density discontinuously jumps and remains $O(1)$ (see fig.~\ref{Re_vs_delta}). 
To aid in explaining the behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in fig.~\ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two. We refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the vegetation density increases, causing the most unstable mode to transition discontinuously.

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 1 asymptotically localizes to the boundary layer near the top of the vegetation, and exhibits an asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold. 
This limit can be understood through asymptotically estimating the sizes of the terms in eq.~\eqref{Orr-somerfield} using $D\sim \delta^{-1}$, $\sigma \sim \delta^{-1}$, and $U\sim \delta$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
The terms balance when $\Rey \sim H/\delta$ (or $\Rey \sim {\Ndg}^{1/2}$) leading to a simplification of eq.~\eqref{Orr-somerfield} as
\begin{align}
(D^2-k^2)^2 \phi- 2 \Rey \Ndg D( SUD\phi) = \sigma \Rey (D^2-k^2)\phi,
\label{eqn:mode1asymp}
\end{align} 
in a region of thickness O($\delta$) near $y=\hg$.
Since the relative magnitude of the terms on the r.h.s. to those on the l.h.s. of this equation is $\Rey/\Ndg^{1/2}$, we expect identical asymptotic behavior for fixed $\Rey/\Ndg^{1/2}$. Therefore the threshold obtained for Mode 1 is $\Rey \propto \Ndg^{1/2}$ (or $\Rey \propto H/\delta$) explaining the numerically observed asymptote (see fig.~\ref{Re_vs_delta}). 
This analysis also concludes that the mode structure is self-similar over the length scale $\delta$ for fixed $\Rey/\Ndg^{1/2}$; the verification of this expectation is shown in fig.~\ref{Asymptotic_mode}, supporting this argument.

On the other hand, the threshold condition for Mode 2 is numerically observed to be $\Rey \propto ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in fig.~\ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of eq.~\eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but fixed $\Rey/\Ndg \sim O(1)$.
Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the vegetation is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
This estimate simplifies eq.~\eqref{Orr-somerfield} to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,  \quad &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, \quad &\text{ for $y>\hg$}.
\end{split}
\label{eqn:mode2asymp}
\end{align}
\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of vegetation density. 
Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The mode shapes approach each other for these small values of $\delta/H$ indicating that the large vegetation density asymptote is reached. Mode 1 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. 
Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. 
The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the vegetation height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}
Since $\Rey/\Ndg$ is the only remaining parameter, the mode shape would converge in the aforementioned limit, in agreement with our numerical results for Mode 2 shown in fig.~\ref{Asymptotic_mode}. 
We interpret Mode 2 as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the top of the vegetation plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).
%On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.

@@

In a real meadow, the drag coefficients are known to vary from bottom to tip of the grass blades \cite{Vivoni98,Nepf00}. 
The eddy viscosity is also known to vary between the flow of unvegetated region and flow through the meadow \cite{Ghisal02}. 
Although an improved analysis of flow with these variation in the model might lead to a better agreement between the observed and the predicted quantities, these variation are not central to the mechanism leading to the instability of flow.
\newline
In conclusion, we note that the threshold flow condition observed in the field and in lab experiments arises due to the presence of vegetation drag. 
We show the origin of this threshold using a consistent continuum hydrodynamic stability analysis for the shear flow that develops as a result of the vegetation-induced drag. 
Such an analysis not only explains quantitative observations, but also provides an understanding of the hydrodynamic instability mechanism. 
The extent of the structure of Mode 2 throughout the depth of the water column implies an efficient transport of suspended tracers compared to that due to the localized structure of Mode 1 near the top of the meadow. 
Our analysis also informs flow structure formation in many other related scenarios, such as flow over coral reefs, permeable sediments, etc. and therefore is expected to have a wider impact.

@@

\bibliography{Grass}{}
\bibliographystyle{plain}
\end{document}

>>>>

\documentclass[aps,prl,twocolumn,superscriptaddress,sort&compress,10pt]{revtex4-1}  % for review and submission
%\documentclass[aps,twocolumn,floatfix,prl,10pt]{revtex4-1}
%\documentclass[letterpaper,10pt,prl,twocolumn,aps]{revtex4-1}
%\documentclass[aps,preprint,floatfix,prl]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color}
\usepackage{longtable}
\usepackage{array}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\graphicspath{ {images/}{images/Growth_Rate_Cont_Plots/} }
\newcommand{\bn}{{\boldsymbol{\hat{n}}}}
\newcommand{\bt}{{\boldsymbol{\hat{t}}}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand{\del}{\partial}
\newcommand{\hg}{h_g}
\newcommand{\Rey}{\text{R}}
\newcommand{\Ndg}{\tilde{N}_g}
\newcommand{\monami}{\textit{monami}}
\newcommand{\ubl}{U_\text{bl}}
\definecolor{tableShade}{gray}{0.8}

\begin{document}
\title{\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
% \author{L. Mahadevan}
% \affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi}
\affiliation{OIST Graduate University, Okinawa 904-0495, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hole Oceanographic Institution, Woods Hole MA 02543 USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

Therefore, to develop the essential mathematical model, we assume the vegetation to be rigid, and verify the assumption \textit{a posteriori}.

@@

The former arises because, in the case of dense vegetation, the shear stress exerted by the bottom surface is expected to be negligible compared to the vegetation drag, whereas the latter is due to the free interface. 

@@

Because this boundary layer develops from purely local dynamics, independent of the influence of the boundaries, we identify it to be analogous to the free shear layer \cite{Ghisal02} in the previous explanation of \monami.

@@

Solving for $\delta$ and $\ubl$ yields $\delta/H = \ubl/U_0=(\Rey\Ndg)^{-1/3}$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, and $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow. 

@@

The asymptotic regime of a thin boundary layer is expected to hold for $\Rey \Ndg \gtrsim 100$. Using these variables, $U_g/U_0 = (\Rey \Ndg)^{-1/2}$. 

@@

Next we substitute $\bu = (U+\tilde{u}, \tilde{v})$, $p=P+\tilde{p}$ in eq.~\eqref{navier-stokes} and expand to linear order to investigate the evolution of small perturbations $(\tilde{u}, \tilde{v})$. These perturbations obey (dropping the tilde decoration)

@@

\begin{table*}
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.4}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{Equation \eqref{base_equ}} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& $\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Dense grass limit &  & Equation \eqref{eqn:mode1asymp} & Equation \eqref{eqn:mode2asymp}  \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{1/2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto H/\delta$		& $\propto H/\delta$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& ~~yes, near $y=\hg$~~			& ~~no, spans water column~~
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the mechanism of waving and the origin of the threshold, we consider the behavior of the instability as a function of the vegetation drag characterized by $\Ndg$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical vegetation density discontinuously jumps and remains $O(1)$ (see fig.~\ref{Re_vs_delta}). 
To aid in explaining this behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in fig.~\ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two. We refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the vegetation density increases, causing the most unstable mode to transition discontinuously.

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 1 asymptotically localizes to the boundary layer near the top of the vegetation, and exhibits an asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold. 
This limit can be understood through asymptotically estimating the sizes of the terms in eq.~\eqref{Orr-somerfield} using $D\sim H/\delta$, $\sigma \sim H/\delta$, and $\ubl \sim \delta/H$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
The terms balance when $\Rey \sim H/\delta$ (or $\Rey \sim {\Ndg}^{1/2}$) leading to a simplification of eq.~\eqref{Orr-somerfield} as
\begin{align}
(D^2-k^2)^2 \phi- 2 \Rey \Ndg D( SUD\phi) = \sigma \Rey (D^2-k^2)\phi,
\label{eqn:mode1asymp}
\end{align} 
in a region of thickness O($\delta$) near $y=\hg$.
Since the relative magnitude of the terms on the r.h.s. to those on the l.h.s. of this equation is $\Rey/\Ndg^{1/2}$, we expect identical asymptotic behavior for fixed $\Rey/\Ndg^{1/2}$. 
Therefore the threshold obtained for Mode 1 is $\Rey \propto \Ndg^{1/2}$ (or $\Rey \propto H/\delta$) explaining the numerically observed asymptote (see fig.~\ref{Re_vs_delta}). 
This analysis also concludes that the mode structure is self-similar over the length scale $\delta$ for fixed $\Rey/\Ndg^{1/2}$; the verification of this expectation is shown in fig.~\ref{Asymptotic_mode}, supporting this argument.

On the other hand, the threshold condition for Mode 2 is numerically observed to be $\Rey \propto ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in fig.~\ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of eq.~\eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but fixed $\Rey/\Ndg \sim O(1)$.
% Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the vegetation is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
In this limit eq.~\eqref{Orr-somerfield} simplifies to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,  \quad &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, \quad &\text{ for $y>\hg$}.
\end{split}
\label{eqn:mode2asymp}
\end{align}
\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of vegetation density. 
Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The mode shapes approach each other for these small values of $\delta/H$ indicating that the large vegetation density asymptote is reached. Mode 1 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. 
Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. 
The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the vegetation height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}
Since $\Rey/\Ndg$ is the only remaining parameter, the mode shape would converge in the aforementioned limit, in agreement with our numerical results for Mode 2 shown in fig.~\ref{Asymptotic_mode}. 
We interpret Mode 2 as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the top of the vegetation plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).
%On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.

@@

In real meadows, the drag coefficients are known to vary from bottom to tip of the grass blades \cite{Vivoni98,Nepf00}. 
The eddy viscosity is also known to vary between the flow of unvegetated region and flow through the meadow \cite{Ghisal02}. 
Although an improved analysis of flow with these variation in the model might lead to a better agreement between the observed and the predicted quantities, these variation are not central to the mechanism leading to the instability of flow.
\newline
In conclusion, we note that the threshold flow condition observed in the field and in lab experiments arises due to the presence of vegetation drag. 
We show the origin of this threshold using a consistent continuum hydrodynamic stability analysis for the shear flow that develops as a result of the vegetation-induced drag. 
Such an analysis not only explains quantitative observations, but also provides an understanding of the hydrodynamic instability mechanism. 
% The extent of the structure of Mode 2 throughout the depth of the water column implies an efficient transport of suspended tracers compared to that due to the localized structure of Mode 1 near the top of the meadow. 
Our analysis also informs flow structure formation in many other related scenarios, such as flow over coral reefs, permeable sediments, etc. and therefore is expected to have a wider impact.

@@

\bibliography{Grass}{}
% \bibliographystyle{plain}
\bibliographystyle{aipauth4-1}
\end{document}
